Sequential rank agreement methods for comparison of ranked lists

### 2015-04-20 latex/paper.Rnw
Revision a4

\section{Introduction}

Ranked lists occur in many applications of statistics. Regression
methods rank predictor variables according to magnitude of association
with outcome, prediction models rank subjects according to their risk
of an event, and genetic studies rank genes according to their
difference in expression across samples.

A common research question is where to stop, \ie, to decide the
maximal significant rank.  In the three examples this would correspond
to the number of significantly associated predictor variables, the
number of patients at high risk of an event, and the number of genes
that are worth to pursue in further experiments, respectively.

In this article we describe some new breakthrough tools for measuring
agreement across a set of lists which soon will enter the state-of-the
art (FIXME: denne sætning skal laves om). The methods should be useful
whenever there are multiple rankings of the same list. The idea is to
define agreement based on the ranks of the first $k$ elements in each
list.

Several other approaches have proposed to measure ... Of these, 
Kendall's $\tau$ is the most well-known but xxx and yyy have proposed
methods that measure the ..

Rank order statistics

Classical pairwise similarity measures have been based on correlations
as is the case for Kendall’s $\tau$ \citep{kendall1948} and Spearman’s
$\rho$ \citep{Spearman1910}. Both correlations measure the similarity of two
ranked lists on a correlation scale from -1 to 1, where 1 implies that
the lists are completely similar and -1 implies that they are
completely dissimilar, in the case of two lists that translates to one
being the reverse of the other, while 0 correspond to random
lists. Correlation based measures treat discrepancies at different
depths of the lists equally, so that a difference at the end
contributes as much as one in the top of the lists.

Kendall's $\tau$ does not make any distinction between errors
towards the top versus the errors towards the bottom

Several modified versions of these measures has been proposed, such as
weighted $\tau$ by Shieh (1998) where each pair of rankings can be
assigned different weights, and the τap (Yilmaz et al., 2008) ap
refers to average precision, which places higher emphasis on the top
of the lists. Spearman’s ρ use the ranks of the variables for
calculation of the correlation and the use of ranks are also employed
in the M measure of Bar-Ilan et al. where the reciprocal rank
differences are used to calculate the similarity measure.  Rather than
looking at correlations between lists some recent papers use the
inter- section of the lists as the basis for a similarity measure. It
can be argued that intersection is a better measure when looking at
non-conjoint lists, but the simple intersection places equal weights
on all depths of the list. Thus, Fagin et al. (2003) and Webber et
al. (2010) attempt to modify the simple intersection in a way that is
top weighted by using the cu- mulative intersection divided by the
depth as the basis for their similarity measures — in the following
referred to as the average overlap (AO).Webber et al. (2010) define
their rank-biased overlap (RBO) by weighting with a converging series
thus ensuring the top is always weighted higher than the (potentially
indefinite) bottom of the lists. The intersection metric by Fagin et
al. (2003) uses the mean of the overlaps at all depths up to d.  While
it is possible to calculate measures that describe how similar lists
are at a given depth with some of the methods described above, the
interpretation of the measures is unclear and as we will demonstrate
it is not trivial to identify a depth at which the lists are no longer
similar. We will use some terminology from agreement analysis as our 2
method resembles this in some aspects. We will refer to the similar
part of the lists, which is expected at the top if it exists, as
having high agreement and the dissimilar part, where items are ranked
randomly, as having low agreement. As a measure for agreement we will
base this on the limits-of-agreement known from agreement between
quantitative variables (Carstensen, 2011) . Since the
limit-of-agreement essentially is a transformation of the pooled
standard deviation of the ranked items we will use that, since it has
a direct and simple interpretation: at given depth how far do the
items deviate from the mean rank. Thus, twice the standard deviation
is a measure of how far apart in ranks the methods classify the items.
The rest of the paper is structured as follows, in the methods section
we show how to calculate and apply the limits-of-agreement-based
measure, sd, to a small example dataset and we perform a comparison to
average overlap. We also present and discuss how changepoint methods
can be applied to identify where a set of lists goes from high
agreement to low agreement. The simulation section shows the
performance of the measure under a number of different scenarios
while the application shows calculations of agreement and average
overlap on a publicly available colorectal cancer dataset. Finally the
findings are discussed along with possible extensions.

The manuscript is organized as follows:

>>>>

\section{Introduction}
Ranking of items or results is common in scientific research and
ranked lists occur as the result of many applications in
statistics. Regression methods rank predictor variables according to
magnitude of association with outcome, prediction models rank subjects
according to their risk of an event, and genetic studies rank genes
according to their difference in gene expression levels across
samples.

When several ranked lists are available a common research question is
to what extent the lists agree on the ranking. In other words: is it
possible to decide on a optimal rank until which the lists agree on
the items?  A typical situation would arise in high-dimensional
genomics studies where several analysis methods had been applied to a
dataset or where a literature search provided differing tables of,
say, top 10 lists of the most important genes associated to an
outcome.  Similarly, this would correspond to the number of associated
predictor variables that are found consistently over the lists, the
number of patients at high risk of an event in all lists, and the
number of genes that are worth to pursue in further experiments in the
three examples above, respectively.

In this article we propose a method to measure agreement among ranked
lists. The proposed method allows for multiple lists simultaneously,
provides a dynamic measure of agreement for each depth in the lists,
places more weight on the top of the list, accommodates censored/incomplete
lists, and has a natural interpretation that directly relates to the
ranks. The general idea is to define agreement based on the ranks of
the first $k$ elements in each list.

The analysis of rankings is hardly new, and several other approaches
have proposed a measure of the ``distance'' between two ranked
lists. Of these, Kendall's $\tau$ \citep{kendall1948} and Spearman’s
$\rho$ \citep{Spearman1910} are the most well-known. Both correlations
measure the similarity of two ranked lists on a correlation scale from
-1 to 1, but neither of them make any distinction between errors
towards the top versus the errors towards the bottom of the lists.
Several modified versions of Kendall's $\tau$ and Spearman's $\rho$
have been proposed. \citet{Shieh1998} proposed a weighted $\tau$ where
each pair of rankings can be assigned different weights, and
\citet{Yilmaz2008} proposed the $\tau_{ap}$ which places higher
emphasis on the top of the lists.
%
Spearman’s $\rho$ use the ranks of the variables for calculation of
the correlation and the use of ranks are also employed in the $M$
measure of \citet{Bar-Ilan2006} where the reciprocal rank differences
are used to calculate the similarity measure.

Recent approaches consider the intersection of the lists as the basis
for a similarity measure

\citetCarterette2009}

Rather than looking at correlations between lists some recent papers
use the intersection of the lists as the basis for a similarity
measure. It can be argued that intersection is a better measure when
looking at non-conjoint lists, but the simple intersection places
equal weights on all depths of the list. Thus, \citet{Fagin2003} and
Webber et al. (2010) attempt to modify the simple intersection in a
way that is top weighted by using the cumulative intersection divided
by the depth as the basis for their similarity measures — in the
following referred to as the average overlap (AO). Webber et
al. (2010) define their rank-biased overlap (RBO) by weighting with a
converging series thus ensuring the top is always weighted higher than
the (potentially indefinite) bottom of the lists. The intersection
metric by Fagin et al. (2003) uses the mean of the overlaps at all
depths up to d.  While it is possible to calculate measures that
describe how similar lists are at a given depth with some of the
methods described above, the interpretation of the measures is unclear
and as we will demonstrate it is not trivial to identify a depth at
which the lists are no longer similar. We will use some terminology
from agreement analysis as our 2 method resembles this in some
aspects. We will refer to the similar part of the lists, which is
expected at the top if it exists, as having high agreement and the
dissimilar part, where items are ranked randomly, as having low
agreement. As a measure for agreement we will base this on the
limits-of-agreement known from agreement between quantitative
variables (Carstensen, 2011) . Since the limit-of-agreement
essentially is a transformation of the pooled standard deviation of
the ranked items we will use that, since it has a direct and simple
interpretation: at given depth how far do the items deviate from the
mean rank. Thus, twice the standard deviation is a measure of how far
apart in ranks the methods classify the items.  The rest of the paper
is structured as follows, in the methods section we show how to
calculate and apply the limits-of-agreement-based measure, sd, to a
small example dataset and we perform a comparison to average
overlap. We also present and discuss how changepoint methods can be
applied to identify where a set of lists goes from high agreement to
low agreement. The simulation section shows the performance of the
measure under a number of different scenarios while the application
shows calculations of agreement and average overlap on a publicly
available colorectal cancer dataset. Finally the findings are
discussed along with possible extensions.

The manuscript is organized as follows: In the next section we will
define ...
